---
title: "CUNY Spring 2021 Data 622 HW #2"
author: "Adam Rich"
output:
  html_document: 
    theme: "cerulean"
editor_options:
  chunk_output_type: "console"
---


```{r include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE)
suppressPackageStartupMessages({
  library(alrtools, quietly = TRUE, warn.conflicts = FALSE)
  library(knitr, quietly = TRUE, warn.conflicts = FALSE)
  library(tidyverse, quietly = TRUE, warn.conflicts = FALSE)
  library(tree, quietly = TRUE, warn.conflicts = FALSE)
  library(caret, quietly = TRUE, warn.conflicts = FALSE)
  library(pROC, quietly = TRUE, warn.conflicts = FALSE)
  library(broom, quietly = TRUE, warn.conflicts = FALSE)
  library(faraway, quietly = TRUE, warn.conflicts = FALSE)
  library(palmerpenguins, quietly = TRUE, warn.conflicts = FALSE)
  library(GGally, quietly = TRUE, warn.conflicts = FALSE)
})
```



```{r include=FALSE}
source('appendix.R')
```


Homework # 2 (Generative Models) (100 points)
Due on March 12, 11:59pm EST
We will be working with the Penguin dataset again as we did for Homework #1.
Please use “Species” as your target variable. For this assignment, you may want to
drop/ignore the variable “year”.
Using the target variable, Species, please conduct:
a. Linear Discriminant Analysis (30 points):
a. You want to evaluate all the ‘features’ or dependent variables and see
what should be in your model. Please comment on your choices.
b. Just a suggestion: You might want to consider exploring featurePlot
on the caret package. Basically, you look at each of the
features/dependent variables and see how they are different based on
species. Simply eye-balling this might give you an idea about which
would be strong ‘classifiers’ (aka predictors).
c. Fit your LDA model using whatever predictor variables you deem
appropriate. Feel free to split the data into training and test sets
before fitting the model.
d. Look at the fit statistics/ accuracy rates.
b. Quadratic Discriminant Analysis (30 points)
a. Same steps as above to consider
c. Naïve Bayes (30 points)
a. Same steps as above to consider
d. Comment on the models fits/strength/weakness/accuracy for all these three
models that you worked with. (10 points)






## Linear Discriminant Analysis

```{r}
penguins <- palmerpenguins::penguins %>% 
  select(-year)
str(penguins)

caret::featurePlot(
  x = (penguins %>% select(-species)), 
  y = penguins$species)

```


How does one do LDA in R?
What were we taught?














## Appendix

### Packages

```{r}
search()
```

### Code

```{r echo=FALSE, results='asis'}
kode <- read_file('appendix.R')
kode <- gsub('\n##[ ]', '\n', kode)
cat("```\n")
cat(kode)
cat("\n")
cat("```")
```

