---
title: "Data 608-01 Spring 2021: Homework #1"
author: "Adam Rich"
date: "February 14, 2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 4.5, 
  fig.height = 6)

suppressPackageStartupMessages(suppressWarnings(
  library(tidyverse, warn.conflicts = FALSE, quietly = TRUE)
))
```


## Additional Summaries

```{r echo=FALSE, results='hide'}
inc_cols <- cols(
  Rank = col_double(),
  Name = col_character(),
  Growth_Rate = col_double(),
  Revenue = col_double(),
  Industry = col_character(),
  Employees = col_double(),
  City = col_character(),
  State = col_character() )

inc <- readr::read_csv('inc5000_data.csv', col_types = inc_cols)
```

```{r eval=FALSE, echo=TRUE}
inc <- readr::read_csv("inc5000_data.csv")
```

```{r}
# Number of unique industries represented
length(unique(inc$Industry))

# Unique States
# 50 + DC + PR
length(unique(inc$State))
table(inc$State)

# Top 5 industries by count
inc %>% 
  group_by(Industry) %>% 
  summarize(Count = n()) %>% 
  arrange(desc(Count)) %>% 
  head(5)

# Top 5 States by count
inc %>% 
  group_by(State) %>% 
  summarize(Count = n()) %>% 
  arrange(desc(Count)) %>% 
  head(5)

# Top 5 companies by revenue
inc %>% 
  mutate(RevenueBillions = Revenue / 1e9) %>% 
  select(Name, RevenueBillions) %>% 
  arrange(desc(RevenueBillions)) %>% 
  head(5)
```






## Question 1

* Show this two different ways
* By count -- makes it easier to compare similar counts over so many levels
* But it makes it hard to find, so having an option to show it alphabetically would also be good


```{r}

# Sorted by count for easier comparison
inc %>%
  mutate(State = factor(
    inc$State,
    levels = names(sort(table(inc$State), decreasing = FALSE)))) %>%
  group_by(State) %>%
  summarize(Count = n()) %>%
  ggplot() +
  aes(x = State, y = Count) +
  geom_col() +
  coord_flip()
```

```{r}

# Sorted alphabetically for easier finding
inc %>% 
  mutate(State = factor(
    inc$State,
    levels = sort(unique(inc$State), decreasing = TRUE))) %>%
  group_by(State) %>% 
  summarize(Count = n()) %>% 
  ggplot() +
  aes(x = State, y = Count) + 
  geom_col() +
  coord_flip()
```





## Question 2

* Sort industries alphabetically
* Use a boxplot to show variability
* Outliers are shown as individual points
* Show x axis on a log-scale because there are some very large companies
* Add the company count per industry to the label


```{r}

# New York is the 3rd most represented State
table(inc$State) %>% sort(decreasing = TRUE) %>% head()

inc_NY <- inc %>% 
  filter(State == 'NY', complete.cases(.))

# Add company counts to labels
counts_by_industry <- inc_NY %>% 
  group_by(Industry) %>% 
  summarize(Count = n())

counts_by_industry$`Industry:N` <- 
  apply(
    X = counts_by_industry[, 1:2], 
    MARGIN = 1, 
    FUN = function(r) {paste0(r, collapse = ':')})

# Show employee counts on LOG scale because outliers are HUGE
inc_NY %>% 
  inner_join(counts_by_industry) %>% 
  mutate(`Industry:N` = factor(
      x = `Industry:N`, 
      levels = sort(unique(`Industry:N`), decreasing = TRUE))) %>% 
  ggplot() +
  aes(x = `Industry:N`, y = Employees) +
  geom_boxplot() +
  scale_y_log10() +
  coord_flip()

```





## Question 3

* Use a boxplot to show variability
* Sort industries by *overall* revenue per employee
* Boxplot shows the variability of the *individual* revenue per employee values


```{r}

ranked <- inc %>% 
  filter(complete.cases(.)) %>% 
  group_by(Industry) %>% 
  summarize(RevPerEE_IndustryWide = sum(Revenue) / sum(Employees)) %>% 
  arrange(RevPerEE_IndustryWide)

# Assuming I should go back to full dataset
inc %>% 
  filter(complete.cases(.)) %>% 
  mutate(
    RevPerEE = Revenue / Employees,
    Industry = factor(Industry, ranked$Industry)) %>% 
  ggplot() +
  aes(x = Industry, y = RevPerEE) +
  geom_boxplot() +
  scale_y_log10() +
  coord_flip()

```




